Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-07-25 analysis/paper/paper.Rmd
Reworking structure and attempting to number sites in overview map

@@

# Combine spatial point data with assemblage data (holds excavation project
# for colouring points in the maps)
sitejoin <- right_join(site_pts, sites, by = c("LOKID" = "askeladden_id"))

# Assemble site map. anc defined first is used for placing the scalebar,
# which required some tweaking to get right.
anc <- as.numeric(c(sitbbox2$ymin, sitbbox2$xmax))
sitemap <-
  ggplot() +
  geom_sf(data = count_reproj2, fill = "grey", colour = NA) +
  geom_sf(data = sitejoin, aes(fill = excavation_project),
          size = 2.25, shape=21, 
          colour = "black") +
  labs(fill = "Excavation project") +
  scale_fill_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                    values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  ggsn::scalebar(data = site_pts, dist = 10, dist_unit = "km", 
                 transform = FALSE, st.size = 4, height = 0.05, 
                 border.size = 0.05, st.dist = 0.07,
                 anchor = c(x = anc[2] - 8000, y = anc[1]) + 4000) +
  coord_sf(xlim = c(sitbbox2[1], sitbbox2[3]), 
           ylim = c(sitbbox2[2], sitbbox2[4]),
         expand = FALSE) +
 theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank(),
                     legend.position = "none") +
  guides(fill = guide_legend(ncol = 2,
                             title.position = "top", title.hjust = 0.5))

@@

@@

sites %>% 
  mutate(strt_bce = reported_start_bc * -1, 
         end_bce = reported_end_bc * -1,
         dating_method = ifelse(dating_method == "shore_typo", "Shoreline/typology", "Radiocarbon")) %>% 
  select(no, site, dating_method, strt_bce, end_bce) %>% 
  gt() %>% 
    cols_label(
    site = "Site name",
    dating_method = "Dating method",
    strt_bce = "Reported start (BCE)",
    end_bce = "Reported end (BCE)",
  ) %>% 
  tab_header(
    title = "Table 1. Analysed sites."
  ) %>% 
  tab_options(heading.title.font.size = 15)
```
# Multivariate analysis of lithic assemblages

The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies --- studies that have employed more informally driven methods. This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful.
However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics from the Palaeolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995; @barton2021].

@@

The first part of the analysis of the lithic data involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. As this above-mentioned Frison effect, several artefact categories have been collapsed for the CA. (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.

In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material. Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [@baxter1994, 71--77, 103].

A note should also be made on the fact that a few variables that are sometimes invoked for the classification of sites in terms of associated mobility patterns are omitted here [e.g. @Bicho2020; @breivik2016a]. For the assemblage data itself this especially pertains to diversity in tool-types [see also @canessa2021], which has been omitted in light of the previously mentioned Frison effect. Some site specific aspects such as number of features has also been disregarded as taphonomic loss is likely to have led to a chronological bias in their preservation. Similarly, the number of activity areas, effectively number of artefact clusters, however defined, has also been disregarded. This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied [@jørgensen2017]. This pertains for example to the impact of bio-turbation in the form of three-throws, which can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jørgensen2017; @darmark2018e].

@@

Figure \@ref(fig:cor) displays the CA using the lithic count data. The general impression from the plots is that a chronological dimension is associated with the patterning in the data. This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, as well as the horseshoe curve or Guttman effect evident in the column plot [see @baxter1994, 119--120]. The earliest sites tend to be located in the upper left corner of plot A, with increasingly younger sites towards the bottom along the second dimension. Although fewer in number, the sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.

@@

resultslong <- resultslong %>% 
  mutate(site_col = ifelse(site %in% sitelabs, site, "other"))

# Call to plot, grouping loess fits by simulation runs
i1 <- ggplot(resultslong, aes(x = age, y = index, colour = site_col)) + 
  geom_smooth(aes(group = sim), method = "loess", fill = "grey", col = "black",
              alpha = 0.05, se = TRUE) +
  geom_point(size = 0.5, alpha = 0.01) +
  scale_color_manual(name = NULL, breaks = c("Langangen Vestgård 1", 
                                             "Hegna vest 4","Gunnarsrød 6b",
                                             "Prestemoen 1",  "Hovland 3",
                                             "Vallermyrene 4"),
                    values = c("Langangen Vestgård 1" = "red",
                               "Hegna vest 4" = "blue",
                               "Gunnarsrød 6b" = "violet",
                               "Prestemoen 1" = "darkorange",
                               "Hovland 3" = "green",
                               "Vallermyrene 4"  = "cyan1", 
                               "other" = "black"),
                     guide = guide_legend(override.aes = list(size = 3,
                                                              alpha = 1))) +
  scale_x_continuous(breaks = c(seq(-9000,-4000, 1000))) +
  labs(title = "A", x = "Cal. BCE", y = "Curation index (All lithics)") +
  theme_classic() +
  theme(legend.position = "none",
        text=element_text(size=16))

i2 <- ggplot(resultslong, aes(x = age, y = index_flint, colour = site_col)) + 
  geom_smooth(aes(group = sim), method = "loess", fill = "grey", col = "black",
              alpha = 0.05, se = TRUE) +
  geom_point(size = 0.5, alpha = 0.01) +
  scale_color_manual(name = NULL, breaks = c("Langangen Vestgård 1", 
                                             "Hegna vest 4","Gunnarsrød 6b",
                                             "Prestemoen 1",  "Hovland 3",
                                             "Vallermyrene 4"),
                    values = c("Langangen Vestgård 1" = "red",
                               "Hegna vest 4" = "blue",
                               "Gunnarsrød 6b" = "violet",
                               "Prestemoen 1" = "darkorange",
                               "Hovland 3" = "green",
                               "Vallermyrene 4"  = "cyan1", 
                               "other" = "black"),
                     guide = guide_legend(override.aes = list(size = 5,
                                                              alpha = 1))) +
  scale_x_continuous(breaks = c(seq(-9000,-4000, 1000))) +
  labs(title = "B", x = "Cal. BCE", y = "Curation index (Flint)") +
  theme_classic() +
  theme(legend.position = "bottom",
        text=element_text(size=16))

@@

While there are certainly nuances in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear relatively robust. It seems clear that there is a marked drop starting some time just before 8000 BCE, which has stabilised around 7000 BCE. This corresponds well with a chronological framework where the Early Mesolithic, or Flake axe phase, is defined as lasting from c. 9200--8200 BCE [@reitan2016]. The beginning of the phase is set to start with the first human occupation in Norway, which is widely held as originating from South-Scandinavian and North-European regions, and which is to be directly reflected by similarities in the artefact inventories [@bjerck2008; @bang-andersen2012; @fuglestvedt2012; @glørstad2016]. Previous research has proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g. @bjerck2008; @breivik2016; @fuglestvedt2012]. This corresponds very well with the findings reported here, where the earliest assemblages are characterised by relatively high and uniform values on the curation index. The transition to the subsequent Middle Mesolithic, or Microlith phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology [@eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian Steppes [@günther2018; @manninen2021]. The radiocarbon record points towards a coinciding population decline in southern Norway around this time [@nielsen2021]. Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is an issue [@solheim2018a; @solheim2020a; @nielsen2021]. In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE. Referring back to the increasing expediency in the curation data between c. 8200 and 7000 BCE, the Microlith phase could thus represent a transitional period where migrating people and new living practices were propagating through societies in south-eastern Norway --- a process that in light of the curation data would have concluded around 7000 BCE.

The Microlith phase is followed by the Pecked adze phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016]. As is evident from both the CA and the curation data, if we disregard Nedre Hobekk 2, the earliest of the assemblages treated here with this kind of compositional profile is site 40, Gunnarsrød 6b, dated to c. `r sites[sites$site == "Gunnarsrød 6b", c("reported_start_bc", "reported_end_bc")][[1]]*-1`--`r sites[sites$site == "Gunnarsrød 6b", c("reported_start_bc", "reported_end_bc")][[2]]*-1` BCE [@carrasco2014]. The curation data remains stable from around 7000 BCE through the next typological transition at c. 5600 BCE, which, following @reitan2016, signifies the onset of the Nøstvet adze phase. While previously defined as having a slightly longer duration, the Nøstvet phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns in south-eastern Norway [e.g. @jaksland2001; @lindblom1984], and has been explicitly linked to an expedient technological organisation [@glørstad2010, 161; @glørstad2011] --- albeit with the term being somewhat vaguely invoked [@eigeland2015, 127--130]. In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016]. The curation index employed here clearly supports this notion, and suggests that the mobility patterns of the Nøstvet phase were well established in preceding periods.

The subsequent Transverse arrowhead phase (c. 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged- and single-edged points [@reitan2016]. It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c. 4300 BCE [@nielsen2021]. The continued stability of the curation index could indicate that these changes are not related to major shifts in land-use and mobility patterns in the material treated here. However, it is also worth highlighting the strained sample size for the later parts of the Mesolithic, which could mean that the effect is simply missed, especially if the signal is weaker than that for the transition from the Early Mesolithic.

>>>>

```
The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c. 9200--8200 BCE [@reitan2016]. The phase is set to start with the first human occupation in Norway---widely held as originating from southern Scandinavian and northern European regions. This is to be directly reflected by similarities in the artefact inventories [@bjerck2008; @bang-andersen2012; @fuglestvedt2012; @glørstad2016]. Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g. @bjerck2008; @breivik2016; @fuglestvedt2012; but see @viken2018]. The transition to the subsequent Middle Mesolithic, or Microlith Phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology [@eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian steppes [@günther2018; @manninen2021]. The radiocarbon record points towards a coinciding population decline in southern Norway around this time [@nielsen2021]. Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is an issue [@solheim2018a; @solheim2020a; @nielsen2021]. In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE. The Microlith phase is followed by the Pecked Adze Phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016]. Following @reitan2016, the next typological transition at c. 5600 BCE signifies the onset of the Nøstvet Adze Phase. While previously defined as having a slightly longer duration, the Nøstvet Phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns in the Mesolithic of south-eastern Norway [e.g. @jaksland2001; @lindblom1984] In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016]. The subsequent Transverse Arrowhead Phase (c. 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged- and single-edged points [@reitan2016]. It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c. 4300 BCE [@nielsen2021].

@@

# Combine spatial point data with assemblage data (holds excavation project
# for colouring points in the maps). Set up labels to aggregate. There might be
# a way to avoid having to do this manually, for example by using coordinates?
sitejoin <- right_join(site_pts, sites, by = c("LOKID" = "askeladden_id")) %>% 
  mutate(no = as.numeric(no)) %>% 
    mutate(no_m = ifelse(no %in% c(1:10, 20) & !duplicated(no %in% c(1:10, 20)),
                         "1-10, 20",
           #ifelse(no %in% c(7, 9:10, 20) & !duplicated(no %in% c(7, 9:10, 20)), "7, 9-10, 20",
           ifelse(no %in% 11:12 & !duplicated(no %in% 11:12), "11-12", 
           ifelse(no %in% 14:17 & !duplicated(no %in% 14:17), "14-17", 
           ifelse(no %in% 21:22 & !duplicated(no %in% 21:22), "21-22", 
           ifelse(no %in% 25:27 & !duplicated(no %in% 25:27), "25-27",
           ifelse(no %in% c(19, 28:29, 23) & 
                !duplicated(no %in% c(19, 28:29, 23)),
                "19, 23, 28-29",
           ifelse(no %in% c(32:33, 35) & 
                !duplicated(no %in% c(32:33, 35)),
                "32-33, 35",
           #ifelse(no %in% 32:33 & !duplicated(no %in% 32:33), "32-33",
           ifelse(no %in% c(24, 34, 37:40)  &
                    !duplicated(no %in% c(24, 34, 37:40)), 
                  "24, 34, 37-40",
            ifelse(no %in% c(42:43, 45:46, 53)  &
                    !duplicated(no %in% c(42:43, 45:46, 53)), 
                  "42-43, 45-46, 53",
           #ifelse(no %in% 42:43 & !duplicated(no %in% 42:43), "42-43",
           #ifelse(no %in% 45:46 & !duplicated(no %in% 45:46), "45-46",
           ifelse(no %in% 48:50 & !duplicated(no %in% 48:50), "48-50",
           ifelse(no %in% c(31, 47, 51:52) & !duplicated(no %in% c(31, 47, 51:52)), "31, 47, 51-52",
           ifelse(no %in% c(13, 30, 36,
                            41, 44, 54), no, "")))))))))))))

# Assemble site map. anc defined first is used for placing the scalebar,
# which required some tweaking to get right.
anc <- as.numeric(c(sitbbox2$ymin, sitbbox2$xmax))
sitemap <-
  ggplot() +
  geom_sf(data = count_reproj2, fill = "grey", colour = NA) +
  geom_sf(data = sitejoin, aes(fill = excavation_project, label = no),
          size = 2.25, shape = 21, 
          colour = "black") +
  ggrepel::geom_text_repel(data = sitejoin, aes(label = no_m,
                                                 geometry = geom),
                                                 stat = "sf_coordinates",
                           box.padding = 0.2, 
                           max.overlaps = getOption("ggrepel.max.overlaps",
                                                    default = 54)) +
  labs(fill = "Excavation project") +
  scale_fill_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                    values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  ggsn::scalebar(data = site_pts, dist = 10, dist_unit = "km", 
                 transform = FALSE, st.size = 4, height = 0.05, 
                 border.size = 0.05, st.dist = 0.07,
                 anchor = c(x = anc[2] - 8000, y = anc[1]) + 4000) +
  coord_sf(xlim = c(sitbbox2[1], sitbbox2[3]), 
           ylim = c(sitbbox2[2], sitbbox2[4]),
         expand = FALSE) +
 theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank(),
                     legend.position = "none") +
  guides(fill = guide_legend(ncol = 2,
                             title.position = "top", title.hjust = 0.5))

@@

sites %>% 
  mutate(strt_bce = reported_start_bc * -1, 
         end_bce = reported_end_bc * -1,
         dating_method = ifelse(dating_method == "shore_typo", "Shoreline/typology", "Radiocarbon")) %>% 
  select(no, site, dating_method, strt_bce, end_bce) %>% 
  gt() %>% 
    cols_label(
    site = "Site name",
    dating_method = "Dating method",
    strt_bce = "Reported start (BCE)",
    end_bce = "Reported end (BCE)",
  ) %>% 
  tab_header(
    title = "Table 1. Analysed sites."
  ) %>% 
  tab_options(heading.title.font.size = 15)
```
# The analysis of lithic assemblages

@@

The aim of the first of part of the analysis is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies of the Mesolithic in southern Norway that have employed more informally driven methods. This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics from the Palaeolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995; @barton2021].

@@

The first part of the analysis of the lithic data involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. As this part of the analysis partially draws on the above-mentioned Frison effect, several artefact categories have been collapsed for the CA. A full overview of the aggregated variables and their constituent parts is provided in the supplementary material. An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.

In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency of blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material. Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [@baxter1994, 71--77, 103].

A note should also be made on the fact that a few variables that are sometimes invoked for the classification of sites in terms of associated mobility patterns are omitted here [e.g. @Bicho2020; @breivik2016a]. For the assemblage data itself this especially pertains to diversity in tool-types [see also @canessa2021], which has been omitted in light of the Frison effect. Some site specific aspects such as number of features has also been disregarded as taphonomic loss is likely to have led to a chronological bias in their preservation. Similarly, the number of activity areas, effectively number of artefact clusters, however defined, has also been disregarded. This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied [@jørgensen2017]. This pertains for example to the impact of bio-turbation in the form of three-throws, which can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jørgensen2017; @darmark2018e].

@@

The general impression from the CA is that a chronological dimension is associated with the patterning in the data (Figure \@ref(fig:cor)). This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, as well as the horseshoe curve or Guttman effect evident in the column plot [see @baxter1994, 119--120]. The earliest sites tend to be located in the upper left corner of plot A, with increasingly younger sites towards the bottom along the second dimension. Although fewer in number, the sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.

@@

resultslong <- resultslong %>% 
  mutate(site_col = ifelse(site %in% sitelabs, site, "Other sites"))

# Call to plot, grouping loess fits by simulation runs
i1 <- ggplot(resultslong, aes(x = age, y = index, colour = site_col)) + 
  geom_smooth(aes(group = sim), method = "loess", fill = "grey", col = "black",
              alpha = 0.05, se = TRUE) +
  geom_point(size = 0.5, alpha = 0.01) +
  scale_color_manual(name = NULL, breaks = c("Langangen Vestgård 1", 
                                             "Hegna vest 4","Gunnarsrød 6b",
                                             "Prestemoen 1",  "Hovland 3",
                                             "Vallermyrene 4", "Other sites"),
                    values = c("Langangen Vestgård 1" = "red",
                               "Hegna vest 4" = "blue",
                               "Gunnarsrød 6b" = "violet",
                               "Prestemoen 1" = "darkorange",
                               "Hovland 3" = "green",
                               "Vallermyrene 4"  = "cyan1", 
                               "Other sites" = "black"),
                     guide = guide_legend(override.aes = list(size = 3,
                                                              alpha = 1))) +
  scale_x_continuous(breaks = c(seq(-9000,-4000, 1000))) +
  labs(title = "A", x = "Cal. BCE", y = "Curation index (All lithics)") +
  theme_classic() +
  theme(legend.position = "none",
        text=element_text(size=16))

i2 <- ggplot(resultslong, aes(x = age, y = index_flint, colour = site_col)) + 
  geom_smooth(aes(group = sim), method = "loess", fill = "grey", col = "black",
              alpha = 0.05, se = TRUE) +
  geom_point(size = 0.5, alpha = 0.01) +
  scale_color_manual(name = NULL, breaks = c("Langangen Vestgård 1", 
                                             "Hegna vest 4","Gunnarsrød 6b",
                                             "Prestemoen 1",  "Hovland 3",
                                             "Vallermyrene 4", "Other sites"),
                    values = c("Langangen Vestgård 1" = "red",
                               "Hegna vest 4" = "blue",
                               "Gunnarsrød 6b" = "violet",
                               "Prestemoen 1" = "darkorange",
                               "Hovland 3" = "green",
                               "Vallermyrene 4"  = "cyan1", 
                               "Other sites" = "black"),
                     guide = guide_legend(override.aes = list(size = 5,
                                                              alpha = 1))) +
  scale_x_continuous(breaks = c(seq(-9000,-4000, 1000))) +
  labs(title = "B", x = "Cal. BCE", y = "Curation index (Flint)") +
  theme_classic() +
  theme(legend.position = "bottom",
        text=element_text(size=16))

@@

While there are certainly nuances in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear relatively robust. The curation index is relatively high and uniform until some time before 8000 BCE. This corresponds well with the view that the Early Mesolithic is characterised by a relatively high and uniform degree of mobility [e.g. @bjerck2008; @breivik2016; @fuglestvedt2012]. This is followed by a marked increase in expedience, which has stabilised by around 7000 BCE. Again, this corresponds well with the employed chronological framework. Referring back to the demographic changes that are to take place around this transition, the Microlith phase could thus represent a period where migrating people and new living practices were propagating through societies in south-eastern Norway---a process that in light of the curation data would have concluded around 7000 BCE.

The curation index then remains stable for the rest of the Mesolithic. This suggests that the transition to mobility patterns traditionally ascribed to the Nøstvet Phase can indeed be traced back to the Microlith Phase [@solheim2016]. The continued stability of the curation index could also indicate that the demographic changes suggested to take place in the Transverse Arrowhead Phase are not related to major shifts in land-use and mobility patterns in the material treated here. However, it is worth highlighting the strained sample size for the later parts of the Mesolithic, which could mean that the effect is simply missed, especially if the signal is weaker than that for the transition from the Early Mesolithic.
